Follow-up Samples

### 2017-03-31 submission/manuscript_outline_20161024.Rmd
final edits maybe

# Adenoma Model 
adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
adn_imp_vars <- read.csv('../data/process/tables/adn_rf_otu_tax.csv', header = T, row.names = 1)
adn_AUC_run_summary <- read.csv('../data/process/tables/adn_Reduced_ROC_model_summary.csv', header = T)
adn_general_summary <- read.csv('../data/process/tables/adn_reduced_auc_summary.csv', header = T, row.names = 1)

@@

**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05) but not carcinoma versus advanced adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_srnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was a trend for increasingly less community similarity, between samples pre- and post-treatment from adenoma to advanced adenoma to carcinoma.  Despite this, no difference was found in the relative abundance of any specific OTU before and after treatment for adenoma, advanced adenoma, or carcinoma groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).  Using Random Forest models to assess whether changes in post-treatment samples were towards a normal community, only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma. The adenoma model used a total of `r length(rownames(adn_imp_vars))` OTUs, the SRN model used a total of `r length(rownames(srn_imp_vars))` OTUs, and the carcinoma model used a total of `r length(rownames(crc_imp_vars))` OTUs.  A total of `r length(common_vars_summary$otu)` OTUs were common to all three models with many classifying to commensal bacteria (e.g. *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).  Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (Pvalue `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

@@

Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes to the microbiome in those with adenoma, advanced adenoma, and carincoma.  First, we assessed differences between samples pre- and post-treatment in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity metrics.  Second, we explored whether models built to classify adenoma, advanced adenoma, or carcinoma versus normal were able to identify specific community members that differed between initial and follow up.  We also used these models to assess whether changes in the community were toward a more normal microbiome.  Finally, we assessed both whether surgery for adenomas and SRN provided larger community changes or whether chemotherapy or radiation provided additive changes to the microbiome over surgical resection. This study aims to provide evidence as to whether the altered CRC microbiome persists or shifts back towards normal after such interventions.

@@

***The Bacterial Community:*** 
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- or post-treatment.  We found that carcinoma patients had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].  Although no significant differences were observed between advanced adenoma and carcinoma there was an increase in the dissimilarity of the samples pre- and post-treatment from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to advanced adenoma (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].  The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), advanced adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).  Interestingly, when pre- and post-treatment samples were compared, regardless of whether they were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was no difference between pre- and post-treatment samples when investigating alpha diversity metrics for adenoma, advanced adenoma, or carcinoma [Table S1].  Additionally, there was also no difference in the relative abundance of any specific OTU between pre- and post-treatment samples for adenoma, advanced adenoma, or carcinoma only [Figure S1].

***Adenoma Model:***
Using normal and adenoma individauls from a separate cohort we created an adenoma model to help assess whether those with adenoma in this cohort changed towards normal.  The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A].  There was a total of `r length(rownames(adn_imp_vars))` OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A]. There was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "adn" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was also no significant decrease in the positive probability of adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2A].

***Advanced Adenoma Model:***
Using normal and advanced adenoma individuals from a separate cohort we created an advanced adenoma model to help assess whether those with advanced adenoma in this cohort changed towards normal.  The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B].  There was a total of `r length(rownames(srn_imp_vars))` OTUs in the advanced adenoma model [Figure S3B].  Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal. Also similar to the adenoma model there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "SRN" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05) and no significant decrease in the positive probability of advanced adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["SRN", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2B].

***Carcinoma Model:***
Using normal and carcinoma individuals from a separate cohort we created a carcinoma model to help assess whether those with carcinoma in this cohort changed towards normal. The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C].  Interestingly, the AUCs improved from adenoma to advanced adenoma to carcinoma [Figure 2].  There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carcinoma model [Figure S3C].  Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* are also important for carcinoma classification [Figure S3C].  Also, like the adenoma and advanced adenoma models there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was a significant decrease in the positive probability of carcinoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or advanced adenoma.  The one individual still positive for carcinoma after treatment had an increase in carcinoma positive proability between their pre- and post-treatment sample [Figure 2C].

***Adenoma, Advanced Adenoma, and Carcinoma Common OTUs:***
Next, to identify which OTUs are important at all stages of disease, we identified common predictive OTUs within the adenoma, advanced adenoma, and carcinoma models.  When we compared the three different models with each other there were a total of `r length(rownames(common_vars_summary))` common OTUs.  Some of the most common taxonomic identifications belonged to *Bacteroides*, *Blautia*, *Anaerostipes*, *Lachnospiraceae*, and *Dorea*.  The vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].

@@

## Discussion
This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differs and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].  Here we show that the microbiome changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide an additive change.  Although some of the important OTUs classified to genera from bacteria considered the "usual suspects"" (e.g. *Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not.  The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, advanced adenoma, and carcinoma models.  This suggests that members within the commensal community may be the first that change during CRC pathogenesis.  These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].

Unlike previous studies on the microbiome and CRC, ours focuses on comparing the microbiome during recovery from treatment in adenoma, advanced adenoma, and carcinoma groups.  Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], the majority of important OTUs taxonmically classified to commensal bacteria [Figure S3].  Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.

@@

A limitation, in our study, was that there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma or advanced adenoma versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and advanced adenoma (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).  These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time. Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1]. However, there are two reasons that this may not be the case.  First, the advanced adenoma group had a higher dissimilarity than adenoma but lower dissimilarity then carcinoma and the collection time to their follow up samples was less than the adenoma group.  Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].

@@

Assessment of the most important OTUs to the model involved counting the number of times an OTU was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different splits run.  This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs. The final collated list of variables was then run through the mtry optimization again. Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.  For all three models the final optimized mtry was `r `ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`

@@

***Statistical Analysis:***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxson ranked sum test.  Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.

@@

>>>>

# Adenoma Model 
adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_model_top_vars_MDA_Summary.csv', header = T)
adn_imp_vars <- read.csv('../data/process/tables/adn_rf_otu_tax.csv', header = T, row.names = 1)
adn_AUC_run_summary <- read.csv('../data/process/tables/adn_Reduced_ROC_model_summary.csv', header = T)
adn_general_summary <- read.csv('../data/process/tables/adn_reduced_auc_summary.csv', header = T, row.names = 1)

@@

**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05) but not carcinoma versus advanced adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_srnVcrc"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was a trend for increasingly less community similarity, between samples pre- and post-treatment from adenoma to advanced adenoma to carcinoma.  Despite this, no difference was found in the relative abundance of any specific OTU before and after treatment for adenoma, advanced adenoma, or carcinoma groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).  Using Random Forest models to assess whether changes in post-treatment samples were towards a normal community, only those with carcinoma had a significant decrease in positive probability (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05); providing further evidence that treatment has the greatest effect in those with carcinoma. The adenoma model used a total of `r length(rownames(adn_imp_vars))` OTUs, the SRN model used a total of `r length(rownames(srn_imp_vars))` OTUs, and the carcinoma model used a total of `r length(rownames(crc_imp_vars))` OTUs.  A total of `r length(common_vars_summary$otu)` OTUs were common to all three models with many classifying to commensal bacteria (e.g. *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).  Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

@@

Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes towards a normal microbiome in those with adenoma, advanced adenoma, and carincoma.  First, we assessed differences between samples pre- and post-treatment in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity metrics.  Second, we explored whether models built to classify adenoma, advanced adenoma, or carcinoma versus normal were able to identify specific community members that differed between pre- and post-treatment samples.  We also used these models to assess whether changes in the community were toward a more normal microbiome.  Finally, we assessed both whether surgery for adenomas and advanced adenomas provided larger community changes and whether chemotherapy or radiation provided additional changes to the microbiome over surgical resection. This study aims to provide evidence as to whether the altered CRC microbiome persists or shifts back towards normal after such interventions.

@@

***The Bacterial Community:*** 
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (n = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (n = `r change_summary["thetayc", "srn_n"]`), or carcinoma (n = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples.  We found that carcinoma patients had a more dissimilar bacterial community between these samples than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A].  Although no significant differences were observed between advanced adenoma and carcinoma there was an increase in the dissimilarity of pre- and post-treatment samples from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to advanced adenoma (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A].  The bacterial community structure pre- post-treatment was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), advanced adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).  Interestingly, when pre- and post-treatment samples were compared, regardless of whether they were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was no difference between pre- and post-treatment samples when investigating alpha diversity metrics for adenoma, advanced adenoma, or carcinoma [Table S1].  Additionally, there was also no difference in the relative abundance of any specific OTU between these samples for adenoma, advanced adenoma, or carcinoma only [Figure S1].

***Adenoma Model:***
Using normal and adenoma individuals from a separate cohort we created an adenoma model to help assess whether those with adenoma in this cohort changed towards normal.  The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A].  There was a total of `r length(rownames(adn_imp_vars))` OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A]. There was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "adn" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was also no significant decrease in the positive probability of adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2A].

***Advanced Adenoma Model:***
Using normal and advanced adenoma individuals from a separate cohort we created an advanced adenoma model to help assess whether those with advanced adenoma in this cohort changed towards normal.  The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B].  There was a total of `r length(rownames(srn_imp_vars))` OTUs in the advanced adenoma model [Figure S3B].  Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal. Also similar to the adenoma model there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "SRN" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05) and no significant decrease in the positive probability of advanced adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["SRN", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2B].

***Carcinoma Model:***
Using normal and carcinoma individuals from a separate cohort we created a carcinoma model to help assess whether those with carcinoma in this cohort changed towards normal. The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C].  Interestingly, the AUCs improved from adenoma to advanced adenoma to carcinoma [Figure 2].  There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carcinoma model [Figure S3C].  Similar to the adenoma and advanced adenoma models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* were also important for carcinoma classification [Figure S3C].  Also, like the adenoma and advanced adenoma models there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05).  There was a significant decrease in the positive probability of carcinoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or advanced adenoma.  The one individual still positive for carcinoma after treatment had an increase in carcinoma positive proability between their pre- and post-treatment sample [Figure 2C].

***Common OTUs:***
Next, to identify which OTUs were important at all stages of disease, we identified common predictive OTUs within the adenoma, advanced adenoma, and carcinoma models.  When we compared the three different models with each other there were a total of `r length(rownames(common_vars_summary))` common OTUs.  Some of the most common taxonomic identifications belonged to *Bacteroides*, *Blautia*, *Anaerostipes*, *Lachnospiraceae*, and *Dorea*.  The vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2]. The rank of these OTUs, based on median mean decrease in accuracy (MDA), varied higly depending on whether they were in the adenoma, advanced adenoma, or carcinoma model [Figure 3].  As an example, *Bacteroides* (OTU85) was ranked `r filter(crc_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ in the carcinoma model but in both the adenoma and advanced adenoma model ranked `r filter(adn_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ and `r filter(srn_varsMDA, otu == "Otu000085") %>% select(rank)`^st^ respectively [Figure 3].

@@

## Discussion
This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differs and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].  Here we show that the microbiome changes towards normal after treatment for carcinoma and that chemotherapy and radiation did not provide any additional change.  Although some of the important OTUs classified to genera from bacteria considered the "usual suspects" (e.g. *Fusobacterium*, *Porphyromonas*, and *Parvimonas*) many did not.  The majority of important OTUs had taxonomic classifications for resident gut microbes and were common for the adenoma, advanced adenoma, and carcinoma models.  This suggests that members within the commensal community may be the first that change during CRC pathogenesis.  These subtle changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].

Unlike previous studies on the microbiome and CRC, ours focuses on comparing the microbiome during recovery from treatment in adenoma, advanced adenoma, and carcinoma groups.  Although there were differences for genera associated with specific bacterium linked with CRC [Figure S3], most of the important OTUs taxonomically classified to commensal bacteria [Figure S3].  Although these changes may be subtle, due to the lack of significant differences in the bacterial community pre- and post-treatment in adenoma and advanced adenoma [Figure 1]; they support the hypothesis that the first members of the community to change and potentially stay changed even after treatment are those that are commensal bacteria.

@@

A limitation in our study was that there was a significant difference in the time elapsed in the collection of the post-treatment sample between adenoma or advanced adenoma versus carcinoma (P-value `r ifelse(time_pvalues["adnVcrc", "bh"] < 0.05 & time_pvalues["srnVcrc", "bh"] < 0.05, invisible("<"), invisible(">"))` 0.05), with time passed being less for adenoma (`r paste(format(time_summary["adenoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["adenoma", "sd_days"], digits = 0))` days) and advanced adenoma (`r paste(format(time_summary["srn", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["srn", "sd_days"], digits = 0))`) than carcinoma (`r paste(format(time_summary["carcinoma", "mean_days"], digits = 0), label = "\u00B1", format(time_summary["carcinoma", "sd_days"], digits = 0))`).  These results would indicate that some of the differences observed between the carcinoma and adenoma groups could be due to differences in collection time. Specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1]. However, there are two reasons that this may not be the case.  First, the advanced adenoma group had a higher dissimilarity than adenoma but lower dissimilarity then carcinoma and the collection time to their follow up samples was less than the adenoma group.  Second, this confounding would not affect the observations where models were used since they were built using a different cohort [Figures 2 & S2-S3].

@@

Assessment of the most important OTUs to the model involved counting the number of times an OTU was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different splits run.  This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs. The final collated list of variables was then run through the mtry optimization again. Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed.  For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`

@@

***Statistical Analysis:***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).  Comparisons between probabilities as well as overall OTU differences between pre- and post-treatment samples utilized a paired Wilcoxson ranked sum test.  Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those that were BH corrected.  Model rank importance was determined by obtining the median MDA from the 100 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.

@@

**Figure 3: Common OTUs Model Rank Based on Median MDA.** For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
